### 2010-10-24 Thesis/Chapter2.tex
- Update 24 October (final) -
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@189 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

\chapter[Background in continuum mechanics for soft-tissue modelling][Background in continuum mechanics]{Background in continuum mechanics for soft-tissue modelling}
\label{chap2}
\begin{shortAbstract}
As seen in the previous chapter, realistic modelling of organ deformation is a challenging research field that opens the door to new clinical applications including: medical training and rehearsal systems, patient-specific planning of surgical procedures and per-operative guidance based on simulation. In all these cases the clinician needs fast updates of the deformation model to obtain a real-time display of the computed deformations. If for medical training devices the haptic feedback from touching organs merely needs to feel real, the accuracy of the information provided to the clinician in the cases of planning or per-operative guidance is crucial. Therefore, a substantial comprehension of the mechanics involved and a knowledge of the physical properties of anatomical structures are both mandatory in our quest to realistically model the deformation of organs. This chapter will start by introducing the main concepts of continuum mechanics that are fundamental to study the mechanical response of organs. It will then present mathematical models able to describe the different mechanical aspects of materials and briefly discuss the mechanical characterisation of tissues.
\end{shortAbstract}

@@

\section{Description of motion} 	\label{chap2:descriptionMotion}
Let us consider a body $\textbf{\textit{B}}$ of known geometry in a three-dimensional Euclidian space $\field{R}^3$. For a given geometry and loading, $\textbf{\textit{B}}$ will undergo a set of macroscopic changes which is called \emph{deformation}. The region of space occupied by the body at a given time $t$ is termed a \emph{configuration}. A change in the configuration of a continuum body results in a displacement. The displacement of a body has two components: a rigid-body displacement and a deformation. A rigid-body displacement consists of a simultaneous translation and rotation of the body without changing its shape or size. In contrast, a change in shape and\slash or size of the body from an initial configuration $\kappa_{0}$ to a new configuration $\kappa$ is called a deformation. This new configuration $\kappa$ may be designated by the \emph{current} or \emph{deformed configuration}. \OFF

@@

	\subsection{Lagrangian description}	
In the Lagrangian description, the position and physical properties of the particles are referred to a reference configuration $\kappa_{R}$, often chosen to be the undeformed configuration $\kappa_{0}$. Consequently, the current coordinates ($\mathbf{x} \in \kappa$) depend on the reference coordinates ($\mathbf{X} \in \kappa_{0}$): \OFF
\begin{equation}
\textbf{x} = \chi(\textbf{X}, t), \quad \chi(\textbf{X}, 0) = \textbf{X},
\end{equation}
and a typical variable $\phi$ defined over the body is expressed in terms of the coordinates $\mathbf{X}$ and time $t$: \OFF
\begin{equation}
\label{chap2:phi}
\phi = \phi(\textbf{X}, t).
\end{equation}
Let us consider a particule $X$ whose position in the reference configuration is $\mathbf{X}$. From \eqref{chap2:phi}, we note that the value of $\phi$ associated with this fixed particle $X$ changes over time. Hence, the Lagrangian description focuses its attention on the particles of the continuous body and it is usually used in solid mechanics.

	\subsection{Eulerian description}
In constrast, the Eulerian description is interested in changes at fixed locations. This time, the motion and $\phi$ are described with respect to the current position ($\mathbf{x} \in \kappa$): \OFF
\begin{equation}
\phi = \phi(\textbf{x}, t), \quad \textbf{X} = \textbf{X}(\textbf{x}, t).
\end{equation}
For a fixed value of $\mathbf{x} \in \kappa$, $\phi(\mathbf{x},t)$ gives the value of $\phi$ associated with the particle occupying the position $\textbf{x} \in \kappa$, which may very well be a different particle for each new time $ t $. Because a change in time $t$ implies that a different value $\phi$ is observed at the same spatial location $\mathbf{x} \in \kappa$, now probably occupied by a different particle, the Eulerian description is focused on a spatial position. This approach is convenient for the study of fluid flow where the kinematic property of greatest interest is the rate at which change is taking place rather than the shape of the body of fluid at a reference time \citep{Spencer80}. Because we are only interested in the study of solid bodies, the Lagrangian description will be used in the rest of the text.

@@

	\subsection{Deformation gradient tensor}
The displacement field tells us how a particle displaces from the reference to the deformed configuration. More interestingly, we would like to know how a material line would deform (stretch and rotation) within this displacement field. Since the length of a material line $d\mathbf{X}$ can change when going to the deformed configuration as well as its orientation, we can say that $d\mathbf{X}$ deforms into $d\mathbf{x}$. We now seek the relation existing between $d\mathbf{x}$ in the deformed configuration and $d\mathbf{X}$ of the reference configuration. \OFF

@@

The three vectors after deformation $d\mathbf{x}^{(1)}$, $d\mathbf{x}^{(2)}$ and $d\mathbf{x}^{(3)}$ can be obtained with:
\begin{equation}
\label{chap2:dxdX}
d\mathbf{x}^{(i)} = \mathbf{F} \, d\mathbf{X}^{(i)}, \quad i = 1, 2, 3.
\end{equation}
The volume of the parallelepiped that we will note $dV$  can be calculated using the triple product between the three vectors:
\begin{align}
dV &= d\mathbf{X}^{(1)} \cdot d\mathbf{X}^{(2)} \times d\mathbf{X}^{(3)} = (\mathbf{\hat{N}}_{1} \cdot \mathbf{\hat{N}}_{2} \times \mathbf{\hat{N}}_{3}) \, dX^{(1)} dX^{(2)} dX^{(3)} \notag \\
&= dX^{(1)} dX^{(2)} dX^{(3)}, \label{chap2:dV}
\end{align}
where $\mathbf{\hat{N}}_{i}$ is the unit vector along $d\mathbf{X}^{i}$. Similarly, we can compute the corresponding volume in the deformed configuration by:
\begin{align}
dv &= d\mathbf{x}^{(1)} \cdot d\mathbf{x}^{(2)} \times d\mathbf{x}^{(3)} \notag \\
&=  (\mathbf{F} \cdot \mathbf{\hat{N}}_{1}) \cdot (\mathbf{F} \cdot \mathbf{\hat{N}}_{2}) \times (\mathbf{F} \cdot \mathbf{\hat{N}}_{3}) \,  dX^{(1)} dX^{(2)} dX^{(3)} \quad \text{by \eqref{chap2:dxdX}} \notag \\
&= \det \mathbf{F} \, dX^{(1)} dX^{(2)} dX^{(3)}. \label{chap2:dv}
\end{align}
The determinant of $\mathbf{F}$ is called the \emph{Jacobian} and it is denoted by $J = \det \mathbf{F}$. Therefore, the relation between the deformed volume $ dv $ and the undeformed volume $ dV $ may be written as: \OFF
\begin{equation}
\label{chap2:dvdV}
dv = J dV.
\end{equation}
It is worth noting that the Jacobian  J has the physical meaning of being the local ratio of current to reference volume of a material volume element.

	\subsection{Change of surface}
For similar reasons, let us find the relation between an element of area in the reference configuration $d\mathbf{A}$ which becomes $d\mathbf{a}$ in the deformed configuration. Considering an arbitrary material line $d\mathbf{L}$ in the reference configuration and noting $d\mathbf{l}$ the same line after deformation, the reference and current volumes are given respectively by:
\begin{align}
d\mathbf{V} &= d\mathbf{L} \cdot d\mathbf{A} \\
\mbox{and } \quad d\mathbf{v} &= d\mathbf{l} \cdot d\mathbf{a}.
\end{align}
Using \eqref{chap2:dvdV} which relates the reference and deformed volumes and recalling that $d\mathbf{l} = \mathbf{F} \, d\mathbf{L}$, we have:
\begin{equation}
J \, d\mathbf{L} \cdot d\mathbf{A} = (\mathbf{F} \, d\mathbf{L}) \cdot d\mathbf{a}.
\end{equation}
Since this expression is valid for any vector $d\mathbf{L}$ and by using the property $\mathbf{a} \cdot \mathbf{b} = \mathbf{a} ^{T} \mathbf{b}$, the sought relation between an element of area in the reference configuration $d\mathbf{A}$ and its corresponding area $d\mathbf{a}$ in the deformed configuration may be expressed as: \OFF
\begin{equation}
\label{chap2:dadA}
d\mathbf{a} = J \, \mathbf{F}^{-T} \, d\mathbf{A}.
\end{equation}
% 
%\begin{figure}[h]
%\begin{center}
%\includegraphics[width=13cm]{chapter2/changeSurface.pdf}
%\end{center}
%\caption{Transformation of a surface element under a deformation mapping. \TODO{CHANGE TO A BETTER FIGURE.}}
%\label{chap2:fig-changeSurface}
%\end{figure}

@@

In the following we make use of a Cartesian coordinate system and we consider an infinitesimal tetrahedron as shown on \fig{chap2:fig-CauchyStress}. \OFF
%
\begin{figure}[ht]
\begin{center}
\includegraphics[width=6cm]{chapter2/CauchyStress.pdf}
\end{center}
\caption{Tetrahedral element in Cartesian coordinates.}
\label{chap2:fig-CauchyStress}
\end{figure}
%
We note $-\mathbf{t}_1$, $-\mathbf{t}_2$, $-\mathbf{t}_3$ and $\mathbf{t}$ the stress vectors in the outwards directions on the faces of the tetrahedron whose areas are respectively $\Delta a_1$, $\Delta a_2$, $\Delta a_3$ and $\Delta a$. For the mass $m$ inside the tetrahedron, we have by Newton's second law:
\begin{equation}
\label{chap2:Newton}
\sum \mathbf{F} = \mathbf{t} \Delta a - \mathbf{t}_1 \Delta a_1 - \mathbf{t}_2 \Delta a_2 - \mathbf{t}_3 \Delta a_3 + \Delta v \mathbf{f} = m \mathbf{a},
\end{equation}
where $\mathbf{f}$ is the force per unit volume acting on the body, $\Delta v$ the volume of the tetrahedron and $\mathbf{a}$ the acceleration.
The areas $\Delta a_1$, $\Delta a_2$ and $\Delta a_3$ being the projections of $\Delta a$, they are related to $\Delta a$ in the Cartesian coordinate system $(\mathbf{e}_1, \mathbf{e}_2, \mathbf{e}_3)$ by:
\begin{equation}
\label{chap2:DeltaAi}
\Delta a_i = (\mathbf{n} \cdot \mathbf{e}_i) \Delta a, \quad i=1, 2, 3.
\end{equation}
Dividing \eqref{chap2:Newton} by $\Delta a$, using \eqref{chap2:DeltaAi} and noting that $\Delta v \slash \Delta a \rightarrow 0$ when the tetrahedron shrinks to a point yields:
\begin{align}
\mathbf{t} &= (\mathbf{n} \cdot \mathbf{e}_1) \mathbf{t}_1 + (\mathbf{n} \cdot \mathbf{e}_2) \mathbf{t}_2 + (\mathbf{n} \cdot \mathbf{e}_3) \mathbf{t}_3 \notag \\
&= \mathbf{n} \cdot (\mathbf{e}_1 \mathbf{t}_1 + \mathbf{e}_2 \mathbf{t}_2 + \mathbf{e}_3 \mathbf{t}_3).
\end{align}

@@

	\subsection{Second Piola-Kirchhoff stress tensor}\label{chap2:PiolaStress2}
Let us consider the stress tensor $\mathbf{S}$ resulting in the force $d\mathbf{F}$ in the undeformed area $d\mathbf{A}$ which corresponds to the force $d\mathbf{f}$ in the deformed area $d\mathbf{a}$. We define this stress tensor $\mathbf{S}$ as the \emph{second Piola-Kirchhoff stress tensor}. \OFF
\begin{equation}
d\mathbf{F} = \mathbf{S} \, d\mathbf{A}.
\end{equation}
In other words, the second Piola-Kirchhoff stress tensor gives the transformed current force per unit undeformed area. Similar to the relationship \eqref{chap2:dxdX} between $d\mathbf{x}$ and $d\mathbf{X}$, the force $d\mathbf{f}$ is related to the force $d\mathbf{F}$ via the deformation gradient tensor:
\begin{align}
d\mathbf{F} &= \mathbf{F}^{-1} \, d\mathbf{f} \notag \\
&= \mathbf{F}^{-1} \, (\mathbf{P} \, d\mathbf{A}) \notag \\
&= \mathbf{S} \, d\mathbf{A}.
\end{align}
Therefore, the second Piola-Kirchhoff stress tensor is related to the first Piola-Kirchhoff stress tensor as follows:
\begin{equation}
\mathbf{S} = \mathbf{F}^{-1} \mathbf{P} = J \, \mathbf{F}^{-1} \, \sigma \, \mathbf{F}^{-T}.
\end{equation}

@@

\section{Constitutive equations}
So far, all the relations that we have established are valid for every continuum. Indeed, the derivations carried out so far made no mention of any material. Therefore, we know that these relations are not sufficient to describe the response of a material to a given loading. Moreover, because the stresses result from the deformation of the material, they may be expressed in terms of some measure of this deformation such as the strain. This relationship, which obviously depends on the type of material under consideration, is what we call a \emph{constitutive equation}. Constitutives equations are an attempt to mathematically characterise the behaviour of materials observed from experimental results. \OFF

@@

	\subsection{Elasticity}	\label{chap2:elasticity}
If under applied loads a material stores but does not dissipate energy, and it returns to its orginal shape when the loads are removed, we call that material \emph{elastic}. The work done by the stress is generally dependent on the deformation path. However, for elastic materials the stress does not depend on the path of deformation and therefore the state of stress in the deformed configuration is determined only by the state of deformation. It means that the stress does not depend on the time taken to achieve the deformation or the rate at which the state of deformation is reached.

In addition, if there exists a \emph{strain energy density} function $U_0(\varepsilon)$ such that 
\begin{equation}
\label{chap2:energy}
\boldsymbol \sigma = \pd{U_0}{\varepsilon},
\end{equation}
the material is said to be hyperelastic. However, for \emph{incompressible} materials the stress tensor is not completely determined by the deformation. The hydrostatic pressure\footnote{pressure exerted by a fluid at equilibrium due to the force of gravity} affects the stress and the above equation is then written as:
\begin{equation}
\label{chap2:energy2}
\boldsymbol \sigma = -p \mathbf{I} + \pd{U_0}{\varepsilon},
\end{equation}
where $p$ is the hydrostatic pressure. In order to develop a mathematical model of an hyperelastic material, $U_0$ is expanded in Taylor's series about $\varepsilon = 0$:
\begin{equation}
U_0 = C_0 + C_{ij}\varepsilon_{ij} + \frac{1}{2!}\hat{C}_{ijkl}\varepsilon{ij}\varepsilon_{kl} + \frac{1}{3!}\hat{C}_{ijklmn}\varepsilon_{ij}\varepsilon_{kl}\varepsilon_{mn} + \ldots , 
\end{equation}
where $C_0$, $C_{ij}$ and $\hat{C}$ are material stiffnesses. The form of $U_0$ may varied for different materials. A material is said to be \emph{linear} if the relationship between strain and stress is also linear (see \fig{chap2:fig-linear}). Based on this definition, $U_0$ is a cubic or higher-order function of the strains for non-linear materials whereas $U_0$ is a quadratic function of strain for linear materials. We will start by discussing linear materials for the case of infinitesimal deformations (sections \ref{chap2:Hooke} to \ref{chap2:isotropic}). Consequently, we will not make any distinction between the various measures of stress and strain and use $\boldsymbol \sigma$ and $\boldsymbol \varepsilon$ for stress and strain tensors respectively. 
%
\begin{figure}[ht]
\centering
\includegraphics[width=7cm]{chapter2/linear.pdf}
\caption {The relationship between strain and stress is often non-linear (blue). However, it may be approximated as a linear relationship for small deformations (red). }
\label{chap2:fig-linear}
\end{figure}

@@

	\subsection{Non-linear materials} \label{chap2:non-linear}
For most materials, the linear relation between stress and strain is only valid for small deformation. Beyond a certain threshold, Hooke's law is no longer valid. Note that it does mean that the material is not elastic, it may well recover all its deformation after removal of the loads. However, materials often exibit an elastic threshold in the non-linear range after which permanent deformation occurs. This range of deformation is qualified of \emph{plastic}.

@@

A non-invasive technique called \emph{elastography} was introduced recently to measure Young's modulus. The first device based on this technique was commercialised by Echosens in 2003 \citep{Bastard09}. The techniques of elastography may be classified in two categories: static and dynamic elastography. In static elastography, a quasi-static constraint is applied onto the surface of the tissue and its deformation is measured. If the constraint is known, the Young's modulus may be obtained via Hooke's law. Dynamic elastography relies on shear waves propagation in the tissue. Shear waves are generated by an acoustic driver and the displacement caused by the waves is determined by ultrasounds, magnetic resonance imaging (MRI) or optical tomography. As an example, magnetic resonance elastography is a quantitative technique which allow the generation of maps describing the shear modulus within the tissues (see \fig{chap2:fig-elastography_liver}). However, elastography is still in development and only a few techniques of elastography are actually used clinically. In the future, elastography might be able to provide a map of the material properties of tissues which may be used as input data for any computational model. This technique would be a step towards patient-specific simulation. As diseased organs are often much stiffer than healthy structures, the integration of these data would greatly increase the fidelity of the simulation.
%
\begin{figure}[ht]
\centering
\includegraphics[width=12cm]{chapter2/elastography_liver.pdf}
\caption {Magnetic resonance elastography of the liver in a healthy volunteer and a patient with cirrhosis. Image courtesy of \cite{Talwalkar07}.}
\label{chap2:fig-elastography_liver}
\end{figure}

Yet, precised characterisation of soft tissue is not sufficient to achieve a good simulation. The development of a simplified or optimised version of the model suitable for real-time computation is also mandatory. As we will see in chapter \ref{chap4} and \ref{chap7}, numerous models are available in the literature to describe physics of objects, from fairly simple and naive approaches to more complex and thorough representations. However, the most common technique to solve equations of continuum mechanics is the finite element method. Moreover, since the two main contributions of this thesis deal with finite element methods formulations, the general principles of this method will be detailed in the next section.

>>>>

\chapter{Background in continuum mechanics for soft-tissue modelling}
\label{chap2}
\begin{shortAbstract}
As seen in the previous chapter, realistic modelling of organ deformation is a challenging research field that opens the door to new clinical applications including: medical training and rehearsal systems, patient-specific planning of surgical procedures and per-operative guidance based on simulation. In all these cases the clinician needs fast updates of the deformation model to obtain a real-time display of the computed deformations. If for medical training devices the haptic feedback from touching organs merely needs to feel real, the accuracy of the information provided to the clinician in the cases of planning or per-operative guidance is crucial. Therefore, a substantial comprehension of the mechanics involved and a knowledge of the physical properties of anatomical structures are both mandatory in our quest to realistically model the deformation of organs. This chapter will start by introducing the main concepts of continuum mechanics that are fundamental to study the mechanical response of organs. It will then present mathematical models able to describe the different mechanical aspects of materials and briefly discuss the mechanical characterisation of tissues.
\end{shortAbstract}

@@

\section{Description of motion} 	\label{chap2:descriptionMotion}
Let us consider a body $\textbf{\textit{B}}$ of known geometry in a three-dimensional Euclidean space $\field{R}^3$. For a given geometry and loading, $\textbf{\textit{B}}$ will undergo a set of macroscopic changes which is called \emph{deformation}. The region of space occupied by the body at a given time $t$ is termed a \emph{configuration}. A change in the configuration of a continuum body results in a displacement. The displacement of a body has two components: a rigid-body displacement and a deformation. A rigid-body displacement consists of a simultaneous translation and rotation of the body without changing its shape or size. In contrast, a change in shape and\slash or size of the body from an initial configuration $\kappa_{0}$ to a new configuration $\kappa$ is called a deformation. This new configuration $\kappa$ may be designated by the \emph{current} or \emph{deformed configuration}.

@@

	\subsection{Lagrangian description}	
In the Lagrangian description, the position and physical properties of the particles are referred to a reference configuration $\kappa_{R}$, often chosen to be the undeformed configuration $\kappa_{0}$. Consequently, the current coordinates ($\mathbf{x} \in \kappa$) depend on the reference coordinates ($\mathbf{X} \in \kappa_{0}$): 
\begin{equation}
\textbf{x} = \chi(\textbf{X}, t), \quad \chi(\textbf{X}, 0) = \textbf{X},
\end{equation}
and a typical variable $\phi$ defined over the body is expressed in terms of the coordinates $\mathbf{X}$ and time $t$: 
\begin{equation}
\label{chap2:phi}
\phi = \phi(\textbf{X}, t).
\end{equation}
Let us consider a particle $X$ whose position in the reference configuration is $\mathbf{X}$. From \eqref{chap2:phi}, we note that the value of $\phi$ associated with this fixed particle $X$ changes over time. Hence, the Lagrangian description focuses its attention on the particles of the continuous body and it is usually used in solid mechanics.

	\subsection{Eulerian description}
In contrast, the Eulerian description is interested in changes at fixed locations. This time, the motion and $\phi$ are described with respect to the current position ($\mathbf{x} \in \kappa$): 
\begin{equation}
\phi = \phi(\textbf{x}, t), \quad \textbf{X} = \textbf{X}(\textbf{x}, t).
\end{equation}
For a fixed value of $\mathbf{x} \in \kappa$, $\phi(\mathbf{x},t)$ gives the value of $\phi$ associated with the particle occupying the position $\textbf{x} \in \kappa$, which may very well be a different particle for each new time $ t $. Because a change in time $t$ implies that a different value $\phi$ is observed at the same spatial location $\mathbf{x} \in \kappa$, now probably occupied by a different particle, the Eulerian description is focused on a spatial position. This approach is convenient for the study of fluid flow where the kinematic property of greatest interest is the rate at which change is taking place rather than the shape of the body of fluid at a reference time \citep{Spencer80}. Because we are only interested in the study of solid bodies, the Lagrangian description will be used in the rest of the text.

@@

	\subsection{Deformation gradient tensor}
The displacement field tells us how a particle displaces from the reference to the deformed configuration. More interestingly, we would like to know how a material line would deform (stretch and rotation) within this displacement field. Since the length of a material line $d\mathbf{X}$ can change when going to the deformed configuration as well as its orientation, we can say that $d\mathbf{X}$ deforms into $d\mathbf{x}$. We now seek the relation existing between $d\mathbf{x}$ in the deformed configuration and $d\mathbf{X}$ of the reference configuration.

@@

The three vectors after deformation $d\mathbf{x}^{(1)}$, $d\mathbf{x}^{(2)}$ and $d\mathbf{x}^{(3)}$ can be obtained with:
\begin{equation}
\label{chap2:dxdX}
d\mathbf{x}^{(i)} = \mathbf{F} \, d\mathbf{X}^{(i)}, \quad i = 1, 2, 3.
\end{equation}
The volume of the parallelepiped that we will note $dV$  can be calculated using the triple product between the three vectors:
\begin{align}
dV &= d\mathbf{X}^{(1)} \cdot d\mathbf{X}^{(2)} \times d\mathbf{X}^{(3)} = (\mathbf{\hat{N}}_{1} \cdot \mathbf{\hat{N}}_{2} \times \mathbf{\hat{N}}_{3}) \, dX^{(1)} dX^{(2)} dX^{(3)} \notag \\
&= dX^{(1)} dX^{(2)} dX^{(3)}, \label{chap2:dV}
\end{align}
where $\mathbf{\hat{N}}_{i}$ is the unit vector along $d\mathbf{X}^{i}$. Similarly, we can compute the corresponding volume in the deformed configuration by:
\begin{align}
dv &= d\mathbf{x}^{(1)} \cdot d\mathbf{x}^{(2)} \times d\mathbf{x}^{(3)} \notag \\
&=  (\mathbf{F} \cdot \mathbf{\hat{N}}_{1}) \cdot (\mathbf{F} \cdot \mathbf{\hat{N}}_{2}) \times (\mathbf{F} \cdot \mathbf{\hat{N}}_{3}) \,  dX^{(1)} dX^{(2)} dX^{(3)} \quad \text{by \eqref{chap2:dxdX}} \notag \\
&= \det \mathbf{F} \, dX^{(1)} dX^{(2)} dX^{(3)}. \label{chap2:dv}
\end{align}
The determinant of $\mathbf{F}$ is called the \emph{Jacobian} and it is denoted by $J = \det \mathbf{F}$. Therefore, the relation between the deformed volume $ dv $ and the undeformed volume $ dV $ may be written as: 
\begin{equation}
\label{chap2:dvdV}
dv = J dV.
\end{equation}
It is worth noting that the Jacobian  J has the physical meaning of being the local ratio of current to reference volume of a material volume element.

	\subsection{Change of surface}
For similar reasons, let us find the relation between an element of area in the reference configuration $d\mathbf{A}$ which becomes $d\mathbf{a}$ in the deformed configuration. Considering an arbitrary material line $d\mathbf{L}$ in the reference configuration and noting $d\mathbf{l}$ the same line after deformation, the reference and current volumes are given respectively by:
\begin{align}
d\mathbf{V} &= d\mathbf{L} \cdot d\mathbf{A} \\
\mbox{and } \quad d\mathbf{v} &= d\mathbf{l} \cdot d\mathbf{a}.
\end{align}
Using \eqref{chap2:dvdV} which relates the reference and deformed volumes and recalling that $d\mathbf{l} = \mathbf{F} \, d\mathbf{L}$, we have:
\begin{equation}
J \, d\mathbf{L} \cdot d\mathbf{A} = (\mathbf{F} \, d\mathbf{L}) \cdot d\mathbf{a}.
\end{equation}
Since this expression is valid for any vector $d\mathbf{L}$ and by using the property $\mathbf{a} \cdot \mathbf{b} = \mathbf{a} ^{T} \mathbf{b}$, the sought relation between an element of area in the reference configuration $d\mathbf{A}$ and its corresponding area $d\mathbf{a}$ in the deformed configuration may be expressed as:
\begin{equation}
\label{chap2:dadA}
d\mathbf{a} = J \, \mathbf{F}^{-T} \, d\mathbf{A}.
\end{equation}
% 
%\begin{figure}[h]
%\begin{center}
%\includegraphics[width=13cm]{chapter2/changeSurface.pdf}
%\end{center}
%\caption{Transformation of a surface element under a deformation mapping. \TODO{CHANGE TO A BETTER FIGURE.}}
%\label{chap2:fig-changeSurface}
%\end{figure}

@@

In the following we make use of a Cartesian coordinate system and we consider an infinitesimal tetrahedron as shown on \fig{chap2:fig-CauchyStress}. 
%
\begin{figure}[ht]
\begin{center}
\includegraphics[width=6cm]{chapter2/CauchyStress.pdf}
\end{center}
\caption{Tetrahedral element in Cartesian coordinates.}
\label{chap2:fig-CauchyStress}
\end{figure}
%
We note $-\mathbf{t}_1$, $-\mathbf{t}_2$, $-\mathbf{t}_3$ and $\mathbf{t}$ the stress vectors in the outwards directions on the faces of the tetrahedron whose areas are respectively $\Delta a_1$, $\Delta a_2$, $\Delta a_3$ and $\Delta a$. For the mass $m$ inside the tetrahedron, we have by Newton's second law:
\begin{equation}
\label{chap2:Newton}
\sum \mathbf{F} = \mathbf{t} \Delta a - \mathbf{t}_1 \Delta a_1 - \mathbf{t}_2 \Delta a_2 - \mathbf{t}_3 \Delta a_3 + \Delta v \mathbf{f} = m \mathbf{a},
\end{equation}
where $\mathbf{f}$ is the force per unit volume acting on the body, $\Delta v$ the volume of the tetrahedron and $\mathbf{a}$ the acceleration.
The areas $\Delta a_1$, $\Delta a_2$ and $\Delta a_3$ being the projections of $\Delta a$, they are related to $\Delta a$ in the Cartesian coordinate system $(\mathbf{e}_1, \mathbf{e}_2, \mathbf{e}_3)$ by:
\begin{equation}
\label{chap2:DeltaAi}
\Delta a_i = (\mathbf{n} \cdot \mathbf{e}_i) \Delta a, \quad i=1, 2, 3.
\end{equation}
Dividing \eqref{chap2:Newton} by $\Delta a$, using \eqref{chap2:DeltaAi} and noting that $\Delta v \slash \Delta a \rightarrow 0$ when the tetrahedron shrinks to a point yields:
\begin{align}
\mathbf{t} &= (\mathbf{n} \cdot \mathbf{e}_1) \mathbf{t}_1 + (\mathbf{n} \cdot \mathbf{e}_2) \mathbf{t}_2 + (\mathbf{n} \cdot \mathbf{e}_3) \mathbf{t}_3 \notag \\
&= \mathbf{n} \cdot (\mathbf{e}_1 \mathbf{t}_1 + \mathbf{e}_2 \mathbf{t}_2 + \mathbf{e}_3 \mathbf{t}_3).
\end{align}

@@

	\subsection{Second Piola-Kirchhoff stress tensor}\label{chap2:PiolaStress2}
Let us consider the stress tensor $\mathbf{S}$ resulting in the force $d\mathbf{F}$ in the undeformed area $d\mathbf{A}$ which corresponds to the force $d\mathbf{f}$ in the deformed area $d\mathbf{a}$. We define this stress tensor $\mathbf{S}$ as the \emph{second Piola-Kirchhoff stress tensor}. 
\begin{equation}
d\mathbf{F} = \mathbf{S} \, d\mathbf{A}.
\end{equation}
In other words, the second Piola-Kirchhoff stress tensor gives the transformed current force per unit undeformed area. Similar to the relationship \eqref{chap2:dxdX} between $d\mathbf{x}$ and $d\mathbf{X}$, the force $d\mathbf{f}$ is related to the force $d\mathbf{F}$ via the deformation gradient tensor:
\begin{align}
d\mathbf{F} &= \mathbf{F}^{-1} \, d\mathbf{f} \notag \\
&= \mathbf{F}^{-1} \, (\mathbf{P} \, d\mathbf{A}) \notag \\
&= \mathbf{S} \, d\mathbf{A}.
\end{align}
Therefore, the second Piola-Kirchhoff stress tensor is related to the first Piola-Kirchhoff stress tensor as follows:
\begin{equation}
\mathbf{S} = \mathbf{F}^{-1} \mathbf{P} = J \, \mathbf{F}^{-1} \, \sigma \, \mathbf{F}^{-T}.
\end{equation}

@@

\section{Constitutive equations}
So far, all the relations that we have established are valid for every continuum. Indeed, the derivations carried out so far made no mention of any material. Therefore, we know that these relations are not sufficient to describe the response of a material to a given loading. Moreover, because the stresses result from the deformation of the material, they may be expressed in terms of some measure of this deformation such as the strain. This relationship, which obviously depends on the type of material under consideration, is what we call a \emph{constitutive equation}. Constitutive equations are an attempt to mathematically characterise the behaviour of materials observed from experimental results.

@@

	\subsection{Elasticity}	\label{chap2:elasticity}
If under applied loads a material stores but does not dissipate energy, and it returns to its original shape when the loads are removed, we call that material \emph{elastic}. The work done by the stress is generally dependent on the deformation path. However, for elastic materials the stress does not depend on the path of deformation and therefore the state of stress in the deformed configuration is determined only by the state of deformation. It means that the stress does not depend on the time taken to achieve the deformation or the rate at which the state of deformation is reached.

In addition, if there exists a \emph{strain energy density} function $U_0(\varepsilon)$ such that 
\begin{equation}
\label{chap2:energy}
\boldsymbol \sigma = \pd{U_0}{\varepsilon},
\end{equation}
the material is said to be hyperelastic. However, for \emph{incompressible} materials the stress tensor is not completely determined by the deformation. The hydrostatic pressure\footnote{pressure exerted by a fluid at equilibrium due to the force of gravity} affects the stress and the above equation is then written as:
\begin{equation}
\label{chap2:energy2}
\boldsymbol \sigma = -p \mathbf{I} + \pd{U_0}{\varepsilon},
\end{equation}
where $p$ is the hydrostatic pressure. In order to develop a mathematical model of an hyperelastic material, $U_0$ is expanded in Taylor's series about $\varepsilon = 0$:
\begin{equation}
U_0 = C_0 + C_{ij}\varepsilon_{ij} + \frac{1}{2!}\hat{C}_{ijkl}\varepsilon{ij}\varepsilon_{kl} + \frac{1}{3!}\hat{C}_{ijklmn}\varepsilon_{ij}\varepsilon_{kl}\varepsilon_{mn} + \ldots , 
\end{equation}
where $C_0$, $C_{ij}$ and $\hat{C}$ are material stiffnesses. The form of $U_0$ may varied for different materials. A material is said to be \emph{linear} if the relationship between strain and stress is also linear (see \fig{chap2:fig-linear}). Based on this definition, $U_0$ is a cubic or higher-order function of the strains for non-linear materials whereas $U_0$ is a quadratic function of strain for linear materials. We will start by discussing linear materials for the case of infinitesimal deformations (sections \ref{chap2:Hooke} to \ref{chap2:isotropic}). Consequently, we will not make any distinction between the various measures of stress and strain and use $\boldsymbol \sigma$ and $\boldsymbol \varepsilon$ for stress and strain tensors respectively. 
%
\begin{figure}[ht]
\centering
\includegraphics[width=7cm]{chapter2/linear.pdf}
\caption[Relationship between strain and stress]{The relationship between strain and stress is often non-linear (blue). However, it may be approximated as a linear relationship for small deformations (red). }
\label{chap2:fig-linear}
\end{figure}

@@

	\subsection{Non-linear materials} \label{chap2:non-linear}
For most materials, the linear relation between stress and strain is only valid for small deformation. Beyond a certain threshold, Hooke's law is no longer valid. Note that it does mean that the material is not elastic, it may well recover all its deformation after removal of the loads. However, materials often exhibit an elastic threshold in the non-linear range after which permanent deformation occurs. This range of deformation is qualified of \emph{plastic}.

@@

A non-invasive technique called \emph{elastography} was introduced recently to measure Young's modulus. The first device based on this technique was commercialised by Echosens in 2003 \citep{Bastard09}. The techniques of elastography may be classified in two categories: static and dynamic elastography. In static elastography, a quasi-static constraint is applied onto the surface of the tissue and its deformation is measured. If the constraint is known, the Young's modulus may be obtained via Hooke's law. Dynamic elastography relies on shear waves propagation in the tissue. Shear waves are generated by an acoustic driver and the displacement caused by the waves is determined by ultrasounds, magnetic resonance imaging (MRI) or optical tomography. As an example, magnetic resonance elastography is a quantitative technique which allow the generation of maps describing the shear modulus within the tissues (see \fig{chap2:fig-elastography_liver}). However, elastography is still in development and only a few techniques of elastography are actually used clinically. In the future, elastography might be able to provide a map of the material properties of tissues which may be used as input data for any computational model. This technique would be a step towards patient-specific simulation. As diseased organs are often much stiffer than healthy structures, the integration of these data would greatly increase the fidelity of the simulation.
%
\begin{figure}[ht]
\centering
\includegraphics[width=12cm]{chapter2/elastography_liver.pdf}
\caption[Magnetic resonance elastography of a liver]{Magnetic resonance elastography of the liver in a healthy volunteer and a patient with cirrhosis. Image courtesy of \cite{Talwalkar07}.}
\label{chap2:fig-elastography_liver}
\end{figure}

Yet, precise characterisation of soft tissue is not sufficient to achieve a good simulation. The development of a simplified or optimised version of the model suitable for real-time computation is also mandatory. As we will see in chapter \ref{chap4} and \ref{chap7}, numerous models are available in the literature to describe physics of objects, from fairly simple and naive approaches to more complex and thorough representations. However, the most common technique to solve equations of continuum mechanics is the finite element method. Moreover, since the two main contributions of this thesis deal with finite element methods formulations, the general principles of this method will be detailed in the next section.
